How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-20 Paper/paper.tex
slightly changed title and fixed use of latex commands (add tilde)

\title{How kids code and how we know it:\\An exploratory study in the Scratch project repository }

@@

In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms Scratch projects in our dataset.

@@

The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default. Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

@@

		Projects utilizing variables & 73,577 & 31.51\% \\
		Projects utilizing lists & 9,358 & 4.01\% \\
		Projects with conditional statements & 92,959 & 39.81\% \\
		Projects with recursive loop statements & 180,210 & 77.18\% \\
		Projects with \texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
		Projects with \texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
		\hline
	\end{tabular}
	\caption{Characteristics in the projects in the dataset}
	\label{tbl-characteristics}
\end{table*}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
	\caption{Histogram of the number of sprites in the analyzed projects}
	\label{fig:size-sprites}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
	\caption{Histogram of the number of scripts in the analyzed projects}
	\label{fig:size-scripts}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
	\caption{Histogram of the lines of code in the analyzed projects}
	\label{fig:size-locs}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1indent}
	\caption{Histogram of the maximum code indentation level in the analyzed projects}
	\label{fig:size-indent}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts analyzed scripts}
	\label{fig:cyclomatic}
\end{figure}

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks. For example, in Figure \ref{fig:ui} the cat sprite contains 2 scripts. We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. The maximum code indentation level is used as a code complexity metric. It was calculated using the JSON array depth? \todo{felienne, how did we measure this?}. This corresponds to code indentation level returning 1 for all first-level blocks and increasing by 2 for block enclosures in loops or conditional statements. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.

@@

The lines of code metric was further analyzed to understand code placement. As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites. An additional 3.1\% of the total lines are found in the stage class. More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects. Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.

As a complexity metric we also used the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}. Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. 29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist.}
	\end{minipage}}

\subsection{Code smells}
dead code
duplicates/clones -graph, number of overlapping positioning, number of dead code duplicates
long method
large class
conditional complexity
Dublicated wait block-same trigger caught by more than 1

@@

\subsubsection{Clones Detection for Scratch programs}
\label{clone_Detection}

We distinguish three different forms of clones: Exact clones within one sprite, exact clones between sprites of one program, clones of `wait' blocks. This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.

In total \nProgramsWithClones~(26.9\% of programs with scripts) exhibit one of the three types of cloning.The most common type of cloning is copying entire scripts between sprites. This occurs in 62.939 programs (25.2\%). In 8.220 programs (3.3\%) wait conditions are cloned and in 2.463 files there is a script cloned within a sprite.

\paragraph{Exact clones within one sprite}
In total, we found \todo{x} cases of exact script clones within one sprite. Upon further inspection, we found that some of them consisted of unconnected blocks, as depicted in Figure \ref{fig:Unconnected_clones}. While this is technically a clone, as this is \emph{dead code} and will never be executed, this could be considered unharmful sketching. Looking only at clones starting with an event block, \todo{...} clones remain over \todo{?} files.

@@

\paragraph{Exact clones between sprites of one program}

\paragraph{Clones of `wait' blocks}
Previous work has established that the cloned `wait' blocks can be harmful when modifying Scratch code, which is the reason why we investigated this particular type of cloning. As outlined in section \todo{?} signals can replaced repeated wait blocks, by catching the condition in one place and broadcasting a signal to all other sprites. This is why we found it interesting to see in how many of the cases where a wait block was cloned, the creator of the program also used signals in their program. This indicates that the concept of signals is known, and hence the duplicate condition is not used because the user is not aware of the alternative. In \todo{...} of the cases the program with the duplicated wait block also contained one or more signals.

@@

\subsubsection{Code Clones Between Sprites}
With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common. In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of \todo{right word? procedures} between scripts, only within them. So in many cases there is no other way to share the functionality than making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.

>>>>

\title{How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository }

@@

In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms~Scratch projects in our dataset.

@@

The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random manual sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default. Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

@@

		Projects utilizing variables & 73,577 & 31.51\% \\
		Projects utilizing lists & 9,358 & 4.01\% \\
		Projects with conditional statements & 92,959 & 39.81\% \\
		Projects with recursive loop statements & 180,210 & 77.18\% \\
		Projects with \texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
		Projects with \texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
		Projects with cloned scripts (>=5 blocks) across sprites & 59,634 & 25.54\% \\
		Projects with cloned scripts (>=5 blocks) within sprites & 23,671 & 10.14\%\\
		Projects with cloned procedures (>=5 blocks) & 4,945 & 2.12\%\\
		Projects with cloned functionality blocks (>=4 blocks) across sprites & 60,554 & 25.93\%\\
		Projects with exact clones (>=5 blocks) across sprites & 27,574 & 11.81\%\\
		Projects with exact clones (>=5 blocks) within sprites & 2,043 & 0.87\%\\
		\hline
	\end{tabular}
	\caption{Characteristics in the projects in the dataset}
	\label{tbl-characteristics}
\end{table*}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
	\caption{Histogram of the number of sprites in the analyzed projects}
	\label{fig:size-sprites}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
	\caption{Histogram of the number of scripts in the analyzed projects}
	\label{fig:size-scripts}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
	\caption{Histogram of the lines of code in the analyzed projects}
	\label{fig:size-locs}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1indent}
	\caption{Histogram of the maximum code indentation level in the analyzed projects}
	\label{fig:size-indent}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
	\label{fig:cyclomatic}
\end{figure}

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks. For example, in Figure \ref{fig:ui} the cat sprite contains two scripts. We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. The maximum code indentation level is used as a code complexity metric. It was calculated using the JSON array depth? \todo{felienne, how did we measure this?}. This corresponds to code indentation level returning 1 for all first-level blocks and increasing by 2 for block enclosures in loops or conditional statements. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.

@@

The lines of code metric was further analyzed to understand code organization. As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites. An additional 3.1\% of the total lines are found in the stage class. More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects. Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.

As a complexity metric we also used the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}. Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. 29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist in 1052 (0.45\% of the total) projects.}
	\end{minipage}}

\subsection{Code smells and bad programming practices}

The duplicated code smell is the first smell that we examine. The first step for our analysis is to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same program. For the identification of clones we do not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal. We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}. The next step in the analysis is to determine the minimum size of the scripts that are considered clones instead of incidentally similar. For this, we examine the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}. Based on this distribution, we opt to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/11cloneslines}
	\caption{Number of cloned scripts of different block sizes across and within sprites}
	\label{fig:cloneslines}
\end{figure}

In total, in the dataset we found 170,532 scripts cloned across sprites in 59,634 (25.54\% of the total non-empty) projects. 726,316 copies of those scripts were found, making each clone being copied an average of 4.26 times. Figure \ref{fig:clonesprojects} plots the distribution of clones across projects. The majority of projects contain up to two cloned scripts; 7.24\% of the projects contain three or more. Figure \ref{fig:clonescopies} plots the number of copies of the identified clones. It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974 \footnote{\todo{link}}. Manual inspection of the mostly copied clones revealed that they were scripts used multiple times to cater for different parameter values, sometimes in multiple sprites.

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/11clonesprojects}
	\caption{Histogram of the number of cloned scripts found in the dataset projects}
	\label{fig:clonesprojects}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/11clonescopies}
	\caption{Histogram of the number of copies of the identified clones}
	\label{fig:clonescopies}
\end{figure}

We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of the projects. These are duplication smells that could be avoided by using procedures, which in Scratch can only be defined for specific sprites. The lack of support for project-wide procedures could be the cause for procedure clones. These were measured to 12,878 (7.55\% of the total) clones, in 2.12\% of the projects.

Exact clones were found in 11.81\% of the total projects. Their total number was 66,750 (39.14\% of the total) clones. Exact clones in the same sprite are cases of clearly redundant code. These were rare, found in only 0.87\% of the projects.

Apart from whole scripts we also examined cases with only the first block being different. In first block is commonly the event block and, by separating this one from the subsequent functionality blocks, we could examine if programmers assign the same functionality to handle different events. Cloned functionality blocks

dead code -unconnected blocks,
previous section: A significant number of the defined procedures (13,036 or 6.3\%) are not called in the projects.
This lack of wait-receive syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).

long method
large class

@@

@@

Scratch programs are found to be particularly susceptible to the duplicate code smell: code clones were found in one fourth of the projects in the dataset.

@@

\subsubsection{Code Clones Between Sprites}
With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common. In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of procedures between scripts, only within them. So in many cases there is no other way to share the functionality than making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.
